---
title: "Extracting tiff-layers"
teaching: 0
exercises: 0
---

::::questions
- How do I extract layers from a tiff-file?
::::


::::objectives
- Demonstrate how to extract layers from tiff-files
::::


Tiff (sometimes tif) is a lossless image file format for storing raster graphics. Originally it was designed as a
common format for manufacturers of scanners. Currently it is popular among photographers as an alternative to lossy
formats like jpg and png. From the start it allowed for storing several images in one file, eg scans in different resolutions or multiple pages of a multi-page document. 

What is relevant here, is the application of storing different images of the same object in the same file. It is a 
popular format for storing satelite imagery, where each layer contains different exposures of the same geographical area. Layers can be split by color, where a layer containing the green channel of an rgb-image, can make it easier to
identify vegetation. 

Layers can also contain different spectral channels, eg infrared in one channel and visible light in another.

All this makes it relevant to be able to split a tiff-file in its individual layers. Several applications supports this (and other manipulation of the files):

- [LibTIFF](https://libtiff.gitlab.io/libtiff)
- [GIMP](https://www.gimp.org)
- [ImageMagick](https://imagemagick.org)

One library for Python supporting this is [Pillow](https://pypi.org/project/pillow/)

In R we use the library `raster`: 

```{r}
library(raster)
```

har vi faktisk brug for andet end raster?

Here is an example from 

![Femlagstiff](../fig/poppendorf_5_band.tif)



Så kan vi indlæse filen:

```{r}
tif <- stack("episodes/fig/poppendorf_5_band.tif")
```


Der er fem lag:

```{r}
raster::nlayers(tif)
```

De har navne:
```{r}
names(tif)
```



Og vi kan plotte dem individuelt:

```{r}
plot(tif$poppendorf_5_band_1)
```

Og de kan gemmes individuel:
```{r eval = FALSE}
writeRaster(tif$poppendorf_5_band_1, "filnavn.tif")
```


## Noter

Filen er grabset fra biblioteket `plainview`, mere specifikt er det et udsnit af
sattelitfotos fra Landsat 8, omkring byen Poppendorf, lidt til højre for Rostock.

Husk også at ikke alle TIFF filer har lag. Så hvis du kun kan trække et enkelt 
lag ud, så kan det skyldes at der kun er et.

De her lagdelte TIFF filer bruges ofte  til at gennem forskellige "kanaler" i billedet.

Det kan være en kanal med landskabet set i IR-spektret. Eller tre kanaler med 
hver af farverne rød, grøn og blå. Vælger man kun at kigge på det grønne lag, 
kan det være lettere at identificere hvor vegetationen er.
library


1	Coastal/Aerosol	0,43–0,45 µm
2	Blå	0,45–0,51 µm
3	Grøn	0,53–0,59 µm
4	Rød	0,64–0,67 µm
5	Nær-infrarød (NIR)	0,85–0,88 µm