---
title: "llm noter"
teaching: 42
exercises: 42
---

::: questions
-   FIXME
:::

::: objectives
-   FIXME
:::

https://aicarpentry.org/lessons/#user-novice

linker til

https://southampton-rsg-training.github.io/research-software-ai-tools/index.html

https://southampton-rsg-training.github.io/coding-with-ai/2-ai-assisted-coding.html

https://southampton-rsg-training.github.io/ai-fundamentals-for-researchers/

https://carpentries-lab.github.io/deep-learning-intro/

https://carpentries-incubator.github.io/software-development-with-genai/

https://carpentries-incubator.github.io/genai-implications-cp/

https://carpentries-incubator.github.io/llms-intro-cp/

https://carpentries-incubator.github.io/ai-carpentry-user-workshop-cp/

https://carpentries-incubator.github.io/coding-with-llms-novice/

https://carpentries-incubator.github.io/genai-implications-novice/rigour_transparency.html#what-is-transparency

det meste af det ser interessant ud!

## Fake data

vi vil gerne kunne uploade data til chatbotten, så den kan levere kode der kan
arbejde på vores data.

Use when:

You need an LLM or collaborator to reproduce your data shape but not see the real values.
You want quick, documented synthetic data for demos, tutorials, bug reports, or tests.
Avoid when:

You need formal differential privacy guarantees (FakeDataR focuses on pragmatic masking and synthesis).
You need exact statistical fidelity for modeling benchmarks. It is not a drop-in replacement for originals.


Men det er ikke (altid) en god ide at uploade data på 300 patienter, eller de data
der danner fundamentet for det projekt du har gang i, der skal gøre dig til millionær. Eller udløse et patent. For... Ja, tech-giganterne er ikke helt pålidelige.

Så. Producer fake data!

Det findes der naturligvis et bibliotek til:

```{r}
library(tidyverse)
library(FakeDataR)

mtcars |>
    rownames_to_column() |> 
    generate_fake_data()
```

Der er nogen detaljer omkring modes der skal studeres lidt mere.

```{r eval = FALSE}
bundle <- mtcars |> 
    llm_bundle(
  n = min(100L, nrow(df)),
  level = "medium",
  formats = c("csv","rds"),
  path = tempdir(), filename = "demo_bundle",
  seed = 42, write_prompt = TRUE, zip = TRUE
)

bundle$zip_path

```

1) Upload the schema first so the model understands data types and categories.
2) Upload the fake data file.
3) Ask questions about structure, analysis steps, and code. Do NOT expect correct statistics from fake values.
4) Apply suggested code in your secure environment on the real data.

::: keypoints
-   FIXME

:::