---
title: 'ggplot-animation'
teaching: 10
exercises: 2
---

:::::::::::::::::::::::::::::::::::::: questions 

- How do you write a lesson using R Markdown and `{sandpaper}`?

::::::::::::::::::::::::::::::::::::::::::::::::

::::::::::::::::::::::::::::::::::::: objectives

- Explain how to use markdown with the new lesson template
- Demonstrate how to include pieces of code, figures, and nested challenge blocks

::::::::::::::::::::::::::::::::::::::::::::::::

## Introduction

### Bar race Animation



Vi skal bruge to pakker:
```{r eval = F}
library(ggplot2)
library(gganimate)
```

```{r}
library(gapminder)
```

```{r eval = F}
data <- gapminder %>% 
  filter(continent == "Europe") %>% 
  group_by(year) %>% 
  mutate(rank = rank(gdpPercap)) %>% 
  ungroup()
  

data %>% 
  ggplot()

static_plot <- data %>% 
  ggplot(aes(x=rank, y=gdpPercap)) +
  geom_col() +
  facet_wrap(~year) +
  aes(fill = country) +
  geom_text(col = "gray13",
            hjust = "right",
            aes(label = country),
            y = -5) +
  theme(legend.position = "none") +
  coord_flip() +
  scale_y_reverse() +  
  labs(fill = NULL)
  
static_plot +
  facet_null() +
  aes(group = country) +
  transition_time(year)
static_plot +
  transition_states(year, transition_length = 4, state_length = 1)
```


### animering i alm 

Særligt når vi vil visualisere data der har en 
tidskomponent, kan det være nyttigt at animere data.



Vi skal bruge to pakker:
```{r eval = F}
library(ggplot2)
library(gganimate)
```

```{r}
library(gapminder)
```

```{r eval = F}
data <- gapminder %>% 
  filter(continent == "Europe") %>% 
  group_by(year) %>% 
  mutate(rank = rank(gdpPercap)) %>% 
  ungroup()
  

data %>% 
  ggplot()

static_plot <- data %>% 
  ggplot(aes(x=rank, y=gdpPercap)) +
  geom_col() +
  facet_wrap(~year) +
  aes(fill = country) +
  geom_text(col = "gray13",
            hjust = "right",
            aes(label = country),
            y = -5) +
  theme(legend.position = "none") +
  coord_flip() +
  scale_y_reverse() +  
  labs(fill = NULL)
  
static_plot +
  facet_null() +
  aes(group = country) +
  transition_time(year)
static_plot +
  transition_states(year, transition_length = 4, state_length = 1)
```



::::::::::::::::::::::::::::::::::::: keypoints 

- Use `.md` files for episodes when you want static content
- Use `.Rmd` files for episodes when you need to generate output
- Run `sandpaper::check_lesson()` to identify any issues with your lesson
- Run `sandpaper::build_lesson()` to preview your lesson locally

::::::::::::::::::::::::::::::::::::::::::::::::

