---
title: "Spatial visualization"
teaching: 42
exercises: 42
---

::: questions
-   FIXME
:::

::: objectives
-   FIXME

:::


Visualizing spatial data using the `terra` package can be done using the `plot()` methods for quick and simple results.

If we want to make more advanced visualizations, or if we want to make them more polished and beuautiful, we can use the `gglot2` package, with the help of the `tidyterra` package, augmented by the `ggspatial` package. 

```{r load_libraries}
#| warning: false
#| message: false
library(terra)
library(ggplot2)
library(tidyterra)
library(ggspatial)
```


## Visualization

In `ggplot2` the main functions for making specific graphs are named `geom_*`. Working with spatial raster and vector data, we use `geom_spatraster()` and `geom_spatvector()`.

The rest of the `ggplot2` template, scales, labels etc, are the same.


### Raster Data

We begin by reading ind some data

```{r read_files_demo}
#| eval: false
library(terra)
files <- list.files("data/", pattern = ".tiff", full.names = TRUE)
sentinel <- rast(files)
```

```{r read_files_do}
#| echo: false
#| eval: true
#| message: false
#| warning: false


library(terra)

files <-  c("data/2026-05-25-00_00_2026-05-25-23_59_Sentinel-2_L2A_B02_(Raw).tiff",
"data/2026-05-25-00_00_2026-05-25-23_59_Sentinel-2_L2A_B03_(Raw).tiff",
"data/2026-05-25-00_00_2026-05-25-23_59_Sentinel-2_L2A_B04_(Raw).tiff",
"data/2026-05-25-00_00_2026-05-25-23_59_Sentinel-2_L2A_B08_(Raw).tiff")

sentinel <- rast(files)
```


:::: hint
## Have to download the data again?

Make sure you have a `data` folder in your project folder, and then run this code:

```{r data_download_hint}
#| eval: false
urls <- c("https://raw.githubusercontent.com/KUBDatalab/R-toolbox/main/episodes/data/2026-05-25-00_00_2026-05-25-23_59_Sentinel-2_L2A_B02_(Raw).tiff",
"https://raw.githubusercontent.com/KUBDatalab/R-toolbox/main/episodes/data/2026-05-25-00_00_2026-05-25-23_59_Sentinel-2_L2A_B03_(Raw).tiff",
"https://raw.githubusercontent.com/KUBDatalab/R-toolbox/main/episodes/data/2026-05-25-00_00_2026-05-25-23_59_Sentinel-2_L2A_B04_(Raw).tiff",
"https://raw.githubusercontent.com/KUBDatalab/R-toolbox/main/episodes/data/2026-05-25-00_00_2026-05-25-23_59_Sentinel-2_L2A_B08_(Raw).tiff",
"https://raw.githubusercontent.com/KUBDatalab/R-toolbox/main/episodes/data/cph_neighborhoods_and_frederiksberg.gpkg",
"https://raw.githubusercontent.com/KUBDatalab/R-toolbox/main/episodes/data/parks_cph_frederiksberg.gpkg")

download.file(
  urls,
  destfile = file.path("data", basename(urls)),
  mode = "wb"
)
```

::::

We need to prepare the data a bit; easier to type names, and a rescaling to reflectance values, rather than the 16 bit encoding:



```{r data_prep}
names(sentinel) <- c("blue", "green", "red", "nir")
sentinel <- sentinel/65535
```


Normally we would pipe in the data to the `ggplot()` function, but `geom_spatraster()` requires the data to be specified directly in the function:
```{r initial_spat_plot}
ggplot() +
  geom_spatraster(data  = sentinel, 
  mapping = aes(fill = nir))
```

Note the message. There are about 4.3 million cells in the nir-layer of this dataset. In order to speed up the plot, `geom_spatraster()` resamples the data to about 500,000 cells.

Having the data in the ggplot environment, we immediately gain access to the tools in ggplot2, and it is easy(ish) to adjust colours, titles and the general theming:

```{r nicer_spat_plot}
#| message: false
ggplot() +
  geom_spatraster(data  = sentinel, 
                  mapping = aes(fill = nir)) +
  scale_fill_gradient(low = "black", high = "white") +
  labs(title = "NIR band",
       subtitle = "Copenhagen May 25, 2026",
       fill = "Reflectance") +
  theme_minimal()
```

Best practice in spatial data visualization is to include both an arrow indicating north, and a scale using km as its unit.

```{r better_spat_plot}
#| message: false
ggplot() +
  geom_spatraster(data  = sentinel, 
                  mapping = aes(fill = nir)) +
  scale_fill_gradient(low = "black", high = "white") +
  annotation_north_arrow(location = "tl",
                          pad_x = unit(1, "cm"),
                          pad_y = unit(1, "cm"),
                          style = north_arrow_orienteering(fill = c("white", "red"))) +
    annotation_scale(
    location = "bl", pad_x = unit(1, "cm"), pad_y = unit(1, "cm"),
    line_col = "white", text_col = "white", text_cex = 1, bar_cols = c("red", "white")) +
  labs(title = "NIR band",
       subtitle = "Copenhagen May 25, 2026",
       fill = "Reflectance") +
  theme_minimal()
```

The north arrow is placed in the *t*op*l*eft corner using the `annotation_north_arrow()` function, and padded with 1 cm of extra space in both directions to move it into the map. By default it is black and white, which can be difficult to see on the mostly darkgrey background, and we have changed the colours to red and white.

The scale, added with `annotation_scale()` take similar arguments.

## Facetting

The only part of the standard ggplot2 template not demonstrated yet, is the facet_*`functions.

If we have more than one layer in our SpatRaster object, we can facet, using `lyr` as the faceting variable:

```{r}
ggplot() +
  geom_spatraster(data  = sentinel) +
  facet_wrap(~lyr)
```

### RGB composition

Just like we could make an RGB-composite image using `plotRGB`, we can also make  them using `ggplot`:

```{r}
ggplot() +
  geom_spatraster_rgb(data = sentinel, r = 3, g = 2, b = 1, stretch = "lin")
```

And now adding titles, captions etc works just like it normally does using `ggplot`

:::: challenge
Try to plot the common false color composite, where we place the NIR-band in the red channel, the red band in the green channel and the bu

:::: hint

You can remind yourself of the order of bands in our `sentinel` data running `names(sentinel)`

::::


:::: solution

```{r composite_solution}
#| eval: false
ggplot() +
  geom_spatraster_rgb(data = sentinel, r = 4, g = 3 , b = 2, stretch = "lin")
```

::::
::::

## Vector data

### Load data

We are going to work with the neighbourhood and parks data for Copenhagen and Frederiksberg again:

```{r read_vect_data}
neighbourhoods <- vect("data/cph_neighborhoods_and_frederiksberg.gpkg")
parks <- vect("data/parks_cph_frederiksberg.gpkg")
```


### Choropleth map

If we inspect the `neighbourhoods` data, we can observe that there is a `population` column in the `values` part of it:

```{r inspecting_neighbourhoods}
neighbourhoods
```

We can plot the neighbourhoods, and then colour the polygons, based on the value in `population`;

```{r}
ggplot() +
  geom_spatvector(data = neighbourhoods, aes(fill = population))
```

Note that we map the colour/fill argument to population just like we normally do when we use `ggplot`.

And we can adjust the plot in the exact same way we would do with any other `ggplot`, and add a north-arrow and scale

```{r choropleth}
ggplot() +
  geom_spatvector(data = neighbourhoods, aes(fill = population)) +
  scale_fill_gradient(low = "grey", high = "brown") +
  labs(title = "Population of neighbourhoods in Copenhagen and Frederiksberg", fill = expression("Population")  annotation_north_arrow(location = "tl",
                          pad_x = unit(1, "cm"),
                          pad_y = unit(1, "cm"),
                          style = north_arrow_orienteering(fill = c("white", "red"))) +
    annotation_scale(
    location = "bl", pad_x = unit(1, "cm"), pad_y = unit(1, "cm"),
    line_col = "white", text_col = "white", text_cex = 1, bar_cols = c("red", "white")) +) +
  theme_minimal() +
  theme(
    plot.title = element_text(face = "bold", hjust = 0.5)
  ) +
      annotation_north_arrow(location = "tl",
                          pad_x = unit(1, "cm"),
                          pad_y = unit(1, "cm")) +
    annotation_scale(
    location = "bl", pad_x = unit(1, "cm"), pad_y = unit(1, "cm"), text_cex = 1) 
```

These kinds of maps are useful for visualizing quantities that vary amongst geographical areas. 
The name `choropleth` originates with greek, khōros (χῶρος), meaning "area or region," and plêthos (πλῆθος), meaning "multitude or quantity".

Here we have visualized the population. If we had the data, we could also visualize other quantities: Propotion of elderly people or frequency of certain diseases, to take just two examples.


:::: challenge

With the data we have available, and the techniques we have learned about earlier, try to plot the population density of the neighbourhoods.

:::: hint

Population density is defined as the number of people pr square kilometer (or any other measurement of area). We can find the area of each neighbourhood using the `expanse()` function.
Start by calculating the area, and then divide the population with the area. We can use the `mutate()` just like normally.
::::



:::: solution

```{r pop_dens_plot}
#| eval: false
pop_density <- neighbourhoods |> 
  mutate(area = expanse(neighbourhoods, unit = "km")) |> 
  mutate(pop_density = population/area)

ggplot() +
  geom_spatvector(data = pop_density, aes(fill = pop_density))

```

::::


::::

### Combining layers

If we have more than one layer we would like to plot, a `ggplot` method for that exists as well:

```{r}
ggplot() +
  geom_spatvector(data = neighbourhoods) +
  geom_spatvector(data = parks, fill = "green")
```

We can even add a layer containing raster data, but we will have to think about the order.

:::: challenge

Try to plot the nir-band from `sentinel`, and then the `neighbourhoods` data. 


:::: Hint

Consider playing with the `alpha` and `colour` arguments

::::

:::: solution

One possible way is this.

```{r}
#| eval: false
ggplot() +
  geom_spatraster(data= sentinel, mapping = aes(fill = nir)) +
  geom_spatvector(data = neighbourhoods, alpha = 0, colour= "white") 
```
::::

If `alpha` is not set to 0, the second layer will make the first almost invisible. And the default `colour` for plotting the borders between the neighbourhoods is black and also almost invisible.

::::



### Reprojecting

Handling the mapping of data described as points on a globe to data described as points on a two-dimensional map, is not trivial, but we have access to functions that do it for us.

That is called `reprojecting`. We have data projected in one system, but would like to have it in another.
It is generally important to have all data in the same system, but another advantage is that we can project data described as coordinates in degrees, minutes and seconds into coordinates in km (relative to geodaetic points in the landscape definded by Danish Geodata Agency).

```{r}
ggplot() +
  geom_spatvector(data = neighbourhoods, fill = NA) +
  geom_spatvector(data = parks, fill = "darkgreen", color = NA) +
  coord_sf(
    crs = "EPSG:25832",
    datum = "EPSG:4258"
)
```


Note that the choise of Coordinate Reference System depends on the location. For example it can be relevant to use EPSG::25833 if the data is covering Bornholm.


### Interactive maps

Leaflet



::: keypoints
- FIXME

:::