---
title: "Vector data"
teaching: 42
exercises: 42
---

::: questions
-   FIXME
:::

::: objectives
-   FIXME

:::


```{r}
#| message: false
#| warning: false
#| echo: false

# load the package
library(terra)

# prevent scientific notation
options(scipen = 8)
```

## Introduction

Let's start by getting our hands on some spatial vector data. 

From the website we can download two vector files with the `.gpkg` extension (signifying it is in the format geopackage). The first represent the neighbourhoods of Copenhagen and the municipality of Frederiksberg, the second parks in these two municipalities:

* [Neighborhoods](data/cph_neighborhoods_and_frederiksberg.gpkg)
* [Parks](data/parks_cph_frederiksberg.gpkg)

The first is extracted from the databank of Copenhagen, augmented with the borders of Frederiksberg extracted from OpenStreetMaps using the `osmdata` package.

The second is also extracted from OpenStreetMaps.

Rather than downloading them manually, we can run the following script to download them directly to our `data` folder:

```{r download-files, eval = FALSE}

urls <- c("https://raw.githubusercontent.com/KUBDatalab/R-toolbox/main/episodes/data/cph_neighborhoods_and_frederiksberg.gpkg",
"https://raw.githubusercontent.com/KUBDatalab/R-toolbox/main/episodes/data/parks_cph_frederiksberg.gpkg"

download.file(
  urls,
  destfile = file.path("data", basename(urls)),
  mode = "wb"
)
```

We are going to learn manipulating spatial vector data by analysing the accessibility of parks in the urban environment.

## Read and inspect data

Working with raster data, we used the `raster` function from the `terra` package. Now we are working with vector data, and use the `vec` function:

```{r read_vec_data}
library(terra)

neighborhoods <- vect("data/cph_neighborhoods_and_frederiksberg.gpkg")
parks <- vect("data/parks_cph_frederiksberg.gpkg")
```

If we are not 100% sure of the quality of the data we have downloaded, it is best practice to check that the
geometries are valid. The function `is.valid` checks this, and will return TRUE or FALSE for each object within the data. As the `parks` dataset contains 1014 parks, it will be difficult to identify if one of them is invalid. The `all` function returns TRUE if all values are TRUE:

```{r checking_validity}
#| collapse: true
all(is.valid(parks))
all(is.valid(neighborhoods))
```

We can inspect the data by simply calling the name:

```{r inspect_neighborhoods}
neighborhoods
```

And we can get a closer look at the `name` attribute using the `$` notation, which returns a character vector with the names of the polygons:

```{r neighborhoods_name}
neighborhoods$name
```

:::: challenge
## What about the names of the parks?

Use the same technique to extract the names of the parks.

:::: solution
## Solution

To save space this code only return the first 10 names:
```{r parks_name}
parks$name |> head(n = 10)
```

From this we learn that a lot of what OpenStreetMaps considers parks, do not have a name. For serious work we would probably have to apply some sort of selection to this dataset.

::::
::::

Not all parks have names. We can check for missing values using the `is.na()` function on the `name` attribute. Nesting that into the `table()` function, we can see how many parks are nameless:

```{r missing_park_names}
# Missing park name?
table(is.na(parks$name))
```

Most of what OSM considers parks in this area have no names.

## Simple Plots

For simple visualizations, we use the `plot()` function. We would like to have parks overlaid on the neighborhoods, and must use the `add = TRUE` argument to do that. The first plot is displayed as a base layer, and subsequent plots are added on top of that:

```{r simple_plot}
plot(neighborhoods, main = "Parks in Copenhagen")
plot(parks, add = TRUE, col = "darkgreen")
```

## Polygon Statistics

Both neighborhoods and parks are polygons with an area and a perimeter. 

The information is not stored directly in the data, but can be calculated from the data using the `expance()` and `perim()` functions from the `terra` package:

```{r}
neighborhoods$area_m2 <- expanse(neighborhoods, unit = "m")
neighborhoods$perimeter_m <- perim(neighborhoods)
neighborhoods
```

Note that we do not have to provide a unit for the calculation of the perimeter; the only possible output is in meters.


:::: challenge
## Which neighborhood is the smallest?

Try to figure out which of the neighborhoods (and/or Frederiksberg) is the smallest.


:::: hint
## Hint
We can extract values from the neighborhood object using the `values()` function (just like we did with raster data). After that, we can sort using the tidyverse functions. Remember to run `library(tidyverse)`

An easier solution is to convert the spatial object to a data frame, using either `as.data.frame()` or `as_tibble()`´. The last option give a nicer looking output.

::::


:::: solution
## Solution


```{r neighborhood_size}
#| message: false
#| warning: false
library(tidyverse)
neighborhoods |> 
    as_tibble() |> 
    arrange(area_m2) |> 
    slice(1)
```


::::
::::






:::: challenge
## Add a calculated area and perimeter to the parks

Repeat the calculations we did on the neighborhoods to add areas and perimeters to the parks.

:::: solution
## Solution


```{r parks_area_perim}
#| message: false
#| warning: false
parks$area_m2 <- expanse(parks, unit = "m")
parks$perimeter_m <- perim(parks)

```


::::
::::

```{r}
parks
```

To make further analysis easier, we exclude parks smaller than 2000 $m^2$. We can treat the `park` data as a data frame, and select rows based on a logical comparison of `parks$area_m2` with 2000:

```{r}
parks_clean <- parks[parks$area_m2 > 2000, ] 
```


## Spatial Relationships

Which neighborhoods have the most parks? 

The function `relate()` take two spatVector objects, and return a logical matrix indicating if there is a relation between the objects in the two spatial vectors. It can handle different relations (see the list in the documentation running `?relate`). Here we ask if there is an intersection between the objects - more intuitively: "is there an overlap". That also means that some parks might be located in more than one neighborhood

```{r}
intersect_matrix <- relate(neighborhoods, parks_clean, relation = "intersects") 
```

The result is a logical matrix. We can easily figure out how many parks are in each neighbourhood. Each row is one neighbourhood, each column is a park. And the cell that intersects a row and a column, is TRUE if the park is located (partially) in the neighbourhood. Adding the values in each row, tells us how many parks are in each neighbourhood: 

```{r}
rowSums(intersect_matrix)
```

We do not get the names, but the order of neighbourhoods in the matrix is the same as the order of the names of the neighbourhoods in our `neighbourhoods` data. That means we can augment the result and get a nice table:

```{r}
neighbourhood_stat <- data.frame(name = neighborhoods$name, count = rowSums(intersect_matrix))
neighbourhood_stat
```


## Availability Analysis

## Proximity Analysis

## Exporting Results

::: keypoints
- FIXME

:::