---
title: "Vector data"
teaching: 42
exercises: 42
---

::: questions
-   FIXME
:::

::: objectives
-   FIXME

:::


```{r}
#| message: false
#| warning: false
#| echo: false

# load the package
library("terra")

# prevent scientific notation
options(scipen = 8)
```

## Introduction

Let's start by getting our hands on some spatial vector data. 

From the website we can download two vector files with the `.gpkg` extension (signifying it is in the format geopackage). The first represent the neighbourhoods of Copenhagen and the municipality of Frederiksberg, the second parks in these two municipalities:

* [Neighborhoods](data/cph_neighborhoods_and_frederiksberg.gpkg)
* [Parks](data/parks_cph_frederiksberg.gpkg)

The first is extracted from the databank of Copenhagen, augmented with the borders of Frederiksberg extracted from OpenStreetMaps using the `osmdata` package.

The second is also extracted from OpenStreetMaps.

Rather than downloading them manually, we can run the following script to download them directly to our `data` folder:

```{r download-files, eval = FALSE}

urls <- c("https://raw.githubusercontent.com/KUBDatalab/R-toolbox/main/episodes/data/cph_neighborhoods_and_frederiksberg.gpkg",
"https://raw.githubusercontent.com/KUBDatalab/R-toolbox/main/episodes/data/parks_cph_frederiksberg.gpkg"

download.file(
  urls,
  destfile = file.path("data", basename(urls)),
  mode = "wb"
)
```

We are going to learn manipulating spatial vector data by analysing the accessibility of parks in the urban environment.

## Read and inspect data

Working with raster data, we used the `raster` function from the `terra` package. Now we are working with vector data, and use the `vec` function:

```{r read_vec_data}
library(terra)

neighborhoods <- vect("data/cph_neighborhoods_and_frederiksberg.gpkg")
parks <- vect("data/parks_cph_frederiksberg.gpkg")
```

We can inspect the `name` attribute using the `$` notation, which returns a character vector with the names

OG HER SKAL VI GØRE NOGET, FOR DE ENKELTE POLYGONER HAR IKKE NAVNE HER!

```{r}
parks$name
```

A lot of what is registered as parks in OSM are areas without names.

If we are not 100% sure of the quality of the data we have downloaded, it is best practice to check that the
geometries are valid. The function `is.valid` checks this, and will return TRUE or FALSE for each object within the data. As the `parks` dataset contains 1014 parks, it will be difficult to identify if one of them is invalid. The `all` function returns TRUE if all values are TRUE:

```{r checking_validity}
all(is.valid(parks))
all(is.valid(neighborhoods))
```




::: keypoints
- FIXME

:::