---
title: "Vector data"
teaching: 42
exercises: 42
---

::: questions
-   FIXME
:::

::: objectives
-   FIXME

:::


```{r}
#| message: false
#| warning: false
#| echo: false

# load the package
library("terra")

# prevent scientific notation
options(scipen = 8)
```

## Introduction

Let's start by getting our hands on some spatial vector data. 

From the website we can download two vector files with the `.gpkg` extension (signifying it is in the format geopackage). The first represent the neighbourhoods of Copenhagen and the municipality of Frederiksberg, the second parks in these two municipalities:

* [Neighborhoods](data/cph_neighborhoods_and_frederiksberg.gpkg)
* [Parks](data/parks_cph_frederiksberg.gpkg)

The first is extracted from the databank of Copenhagen, augmented with the borders of Frederiksberg extracted from OpenStreetMaps using the `osmdata` package.

The second is also extracted from OpenStreetMaps.

Rather than downloading them manually, we can run the following script to download them directly to our `data` folder:

```{r download-files, eval = FALSE}

urls <- c("https://raw.githubusercontent.com/KUBDatalab/R-toolbox/main/episodes/data/cph_neighborhoods_and_frederiksberg.gpkg",
"https://raw.githubusercontent.com/KUBDatalab/R-toolbox/main/episodes/data/parks_cph_frederiksberg.gpkg"

download.file(
  urls,
  destfile = file.path("data", basename(urls)),
  mode = "wb"
)
```

We are going to learn manipulating spatial vector data by analysing the accessibility of parks in the urban environment.

## Read and inspect data

Working with raster data, we used the `raster` function from the `terra` package. Now we are working with vector data, and use the `vec` function:

```{r read_vec_data}
library(terra)

neighborhoods <- vect("data/cph_neighborhoods_and_frederiksberg.gpkg")
parks <- vect("data/parks_cph_frederiksberg.gpkg")
```

If we are not 100% sure of the quality of the data we have downloaded, it is best practice to check that the
geometries are valid. The function `is.valid` checks this, and will return TRUE or FALSE for each object within the data. As the `parks` dataset contains 1014 parks, it will be difficult to identify if one of them is invalid. The `all` function returns TRUE if all values are TRUE:

```{r checking_validity}
#| collapse: true
all(is.valid(parks))
all(is.valid(neighborhoods))
```

We can inspect the data by simply calling the name:

```{r inspect_neighborhoods}
neighborhoods
```

And we can get a closer look at the `name` attribute using the `$` notation, which returns a character vector with the names of the polygons:

```{r neighborhoods_name}
neighborhoods$name
```

:::: challenge
## What about the names of the parks?

Use the same technique to extract the names of the parks.

:::: solution
## Solution

To save space this shows only the first 10 names:
```{r parks_name}
parks$name |> head(n = 10)
```

From this we learn that a lot of what OpenStreetMaps considers parks, do not have a name. For serious work we would probably have to apply some sort of selection to this dataset.

::::
::::

Not all parks have names. We can check for missing values using the `is.na()` function on the `name` attribute. Nesting that into the `table()` function, we can see how many parks are nameless:

```{r}
# Missing park name?
table(is.na(parks$name))
```

Most of what OSM considers parks in this area have no names.

## Simple Plots

For simple visualizations, we use the `plot()` function. We would like to have parks overlaid on the neighborhoods, and must use the `add = TRUE` argument to do that. The first plot is displayed as a base layer, and subsequent plots are added on top of that:

```{r}
plot(neighborhoods, main = "Parks in Copenhagen")
plot(parks, add = TRUE, col = "darkgreen")
```

## Polygon Statistics

## Spatial Relationships

## Availability Analysis

## Proximity Analysis

## Exporting Results

::: keypoints
- FIXME

:::