---
title: 'GitHub and R'
teaching: 10
exercises: 2
---

:::::::::::::::::::::::::::::::::::::: questions 

- What actually is github? 
- How is that even relevant for me?

::::::::::::::::::::::::::::::::::::::::::::::::

::::::::::::::::::::::::::::::::::::: objectives

- Learn how to setup git and GitHub in RStudio
- Learn how to use GitHub for backup, sharing, versioning and collaboration

::::::::::::::::::::::::::::::::::::::::::::::::

## What is GitHub?

You might have seen GitHub before. An integral part of writing code is googling
(and chatgpting) for a solution that kinda matches your problem, copying it, and
adjusting the code so it hopefully does what you need it to do.

A lot of the search results you get will go to StackOverflow, but many will end
up on GitHub. And when the chatbot can help you, it is partially because it has 
been trained on GitHub.

Git is a system for keeping track of different versions of code. 
And because code is just text, it works nicely with text as well, as long as it
is formatted in plain text.

You have a file on your computer. You save it, and use git to store _that_ version
in a `git-repo` on your computer. When you make changes, you add (commit) those
changes to the repo, and Git keeps track of all the changes for you. That is 
pretty neat.

What is even better is that you can push _all_ the changes you made to a centralised
online repository, where your local changes will be stored as well. If something
bad happens to your computer, you have a backup, that you can pull down to a 
new computer.


# Interaktion med github i R-sammenhænge.

Der er mange detaljer. Dette er noten om hvordan man får sat sit Personal
Access Token (PAT) i RStudio.

## Update personal access token:

Det gøres lettest ved at bruge pakken `gitcreds`
```{r gitcreds-eksempel, eval = F}
install.packages("gitcreds")
library(gitcreds)
gitcreds_set()
```

Funktionen returnerer:
```{r setcreds-output, eval  =F }
1: Keep these credentials
2: Replace these credentials
3: See the password / token
```

Vælg "2" og paste dit PAT ind på prompten.

Hvor får du så et PAT? Det sker heromkring fra:


Det token får vi heromkring fra:

[https://github.com/settings/tokens](https://github.com/settings/tokens)



::::::::::::::::::::::::::::::::::::: keypoints 

- GitHub is the standard way for writing code collaboratively
- Be careful about data - not everything belongs on GitHub


::::::::::::::::::::::::::::::::::::::::::::::::

