---
title: 'What we forgot to tell you'
teaching: 10
exercises: 2
---

:::::::::::::::::::::::::::::::::::::: questions 

- What are some things we typically do not mention?

::::::::::::::::::::::::::::::::::::::::::::::::

::::::::::::::::::::::::::::::::::::: objectives

- Get an overview of some of the rabbit holes.


::::::::::::::::::::::::::::::::::::::::::::::::

## Introduction

What we forgot to tell you about R.

## Math in R is imprecise!

It might appear that we get a correct result when we ask R to do math.

But we do not. Every time we do math on numbers with decimals, we can expect 
numerical errors. 

This:

```{r num-impres-1}
7/13 - 3/31
```

is different from: 
```{r num-impres-2}
print(7/13 - 3/31, digits = 16)
```
Actually $\frac{7}{13} - \frac{3}{31}$ is equal to $\frac{178}{403}$. If we subtract
one from the other, we should get 0. We do not...

```{r num-impres-3}
(7/13 - 3/31) - (178/403) 
```

Pretty close but not 0.

R, in the best case, uses 64 bits for storing numerics, and that means the best
we can expect to be correct is 15-16 significant digits after the decimal separator.

This means that we have to expect errors, and that they might accumulate.


```{r integrate}
integrand <- function(x) {1/((x+1)*sqrt(x))}
integrate(integrand, lower = 0, upper = Inf)

```


## When should the warning be ignored?

Never. In introductory courses we tend to tell learners that they should not
worry too much about warnings. 

That is because they generally _can_ be ignored, and they tend to get in the
way of learning the ropes.

However. The moment you have learned (a bit of) R, you should stop ignoring 
errors. What you should do is:

1. figure out what the warning is saying
2. figure out why the warning is triggered
3. Figure out the effect on the results of the computation
4. Determine based on step 3, if your results will be erronious

After that, you might decide that the warning can be safely ignored.



## Do not spend too much time optimizing

Uwes maxim:

Computers are cheap and thinking hurts

Uwe Ligges, member of R Core Team

Your first task is to write clear, understandable and correct code.

When you have debugged your code, when it is stable and correct, and easy for
other people to understand, you might consider optimizing it.

But only if that optimization is likely to make a significant difference. 
Spending an hour to save a millisecond is not useful. Unless that function runs
at least 4 million times, in which case you might actually save time.

## Naming stuff is difficult

how to name stuff af jenny bc

## Think global, act local

Or something like that. The scope you are working in matters.


## Write functions

R is a good thing because it is a language.
The power of language is abstraction.
The way to make abstractions in R is to write functions.

og det er når vi begynder at skrive funktoner, at vi bevæger os fra
at bruge R, til at programmere i R.

Når der skrives funktioner - så skriv små funktioner. Der gør en ting. Og kun
en ting. Det er langt lettere at finde fejlen, når (ikke hvis) den opstår.
Og små funktioner har større sandsynlighed for at være nyttige i andre 
sammenhænge.

Funktioner hvis output er en funktion.

```{r func-making-func}
mycumfun <- ecdf(rnorm(10))
mycumfun(0)
```


https://rstudio.github.io/renv/

xkcd.com/1179






 environments.

## Factor variable trap

Når du konverterer faktorer/kategoriske variable - til numeric, så konverter
faktoren til character først:

fin$profit <- as.numeric(as.character(x))
hvor x er en faktor.

Konverterer du en faktor direkte til en numeric, får du faktor levels (1,2,3 etc)
i stedet for de numeriske værdier der egenligt ligger i faktoren.


::::::::::::::::::::::::::::::::::::: keypoints 

- Use `.md` files for episodes when you want static content


::::::::::::::::::::::::::::::::::::::::::::::::

