---
title: 'Principal Component Analysis'
teaching: 10
exercises: 2
---

:::::::::::::::::::::::::::::::::::::: questions 

- How do you write a lesson using R Markdown and `{sandpaper}`?

::::::::::::::::::::::::::::::::::::::::::::::::

::::::::::::::::::::::::::::::::::::: objectives

- Explain how to use markdown with the new lesson template
- Demonstrate how to include pieces of code, figures, and nested challenge blocks

::::::::::::::::::::::::::::::::::::::::::::::::

CAVE! PINGVINERNE ER RET SEPAREREDE ALLEREDE...
FIND ET ANDET DATASÆT

Der er et vin-datasæt, hvor vi desværre ikke ved hvilke druesorter der er
tale om. Men som er bedre.

https://archive.ics.uci.edu/dataset/109/wine

```{r, include = FALSE}
library(tidyverse)
library(palmerpenguins)
library(GGally)
library(ggExtra)
library(patchwork)
penguins <- penguins %>% filter(complete.cases(.))
```

# Principal Component Analyse

PCA is a method for reducing dimensions of a dataset. 

Let us observe some penguins:

```{r looking-at-penguins}
penguins %>% 
  head()
```

There are 8 variables, or dimensions in this dataset. The dimension `year` is 
not one we have any use for here.

Looking just at the numerical variables, there is a lot of variation, just not
variation that we nessecarily can use for predicting stuff:
```{r}
penguins %>% 
  ggplot(aes(bill_length_mm, body_mass_g, color = sex)) +
  geom_point() +
  ggtitle("weight and bill length is not usefull for predicting sex")
  
```

By applying PCA, we can reduce the four dimensions to fewer, that might be 
better at finding patterns.

We do that by making new variables, or dimensions, that are combinations of the
original dimensions. We might make one that is the sum of the four dimensions, 
where we multiply them by 0.1, 4.2, 47 and 0.001 respectively. And another 
where we multiply with some other values. If we choose those values correctly,
we might get a plot that nicely separates the two sexes.

The assumption is that the variation in the numeric data is dependent on the
sex of the penguins. But rather than figuring out how the variation in weight
describes that, we construct a set of dimensions, that describe the over all 
variation in the numeric data.

And rather than doing it by hand, we can get R to do it for us.









så trækker vi oplysninger om pingvin arten ud til senere brug:
```{r species-pingvin, echo  =F}
species_penguins <- clean_penguins %>% pull(species)
species_penguins %>% head()
```

Og kører PCA på den. Det sker med funktionen `prcomp`:
```{r run-pca, echo  =F}
data_penguins <- clean_penguins %>% select(-species)
pca_penguins <- prcomp(data_penguins, scale. = T)
```

Og så kan vi plotte det. Vi får en del med i modellen, men lad os starte
med at se på de principiale komponenter. De ligger i pca_penguins$x.
Vi tilføjer oplysninger om pingvin-arten, og farvelægger efter netop den:
```{r pca-plot}
pca_penguins$x %>% 
  as_tibble() %>% 
  mutate(species = species_penguins) %>% 
  ggplot(aes(PC1, PC2, color = species)) + 
  geom_point()

```
Vi kan sammenligne med et plot af dimensionerne på deres næb:
```{r}
penguins %>% 
  ggplot(aes(bill_length_mm, bill_depth_mm, color = species)) +
  geom_point()
```

Når vi sammenligner, kan vi se at Æselpingvinerne (Gentoo) er pænt
adskilt fra de to andre pingvinarter i plottet.

Det kan vi så bruge. Vej pingvinerne, mål deres næb og deres vinger. Og du
kan afgøre om de er en æselpingvin. 

Ja, det er nok lettere at kigge på pingvinen, og se om den ligner en 
æselpingvin eller ej.

## Hvad ligger der ellers i modellen?

PCA-modellen for pingvinerne hedder pca_penguins. 

Den indeholder flere ting. I x ligger de nye værdier for alle 
pingvinerne i de nye dimensioner, kaldet Principiale Komponenter:
```{r}
pca_penguins$x %>% head()
```


Hvad havde den første pingvin af data?
```{r}
data_penguins %>% slice(1)
```

Og hvor blev den placeret i de nye koordinater?
```{r}
pca_penguins$x[1,]
```
Hvordan fandt vi dem?
```{r}
pca_penguins$rotation
```

```{r}
(43.99279-39.1)* 0.4537532 - (17.16486-18.7)*0.3990472 + (200.96697-181)*0.5768250+ (4207.05706-3750)*0.5496747
```
```{r}

```


::::::::::::::::::::::::::::::::::::: keypoints 

- Use `.md` files for episodes when you want static content
- Use `.Rmd` files for episodes when you need to generate output
- Run `sandpaper::check_lesson()` to identify any issues with your lesson
- Run `sandpaper::build_lesson()` to preview your lesson locally

::::::::::::::::::::::::::::::::::::::::::::::::

