---
title: 'Table One'
teaching: 10
exercises: 2
---

:::::::::::::::::::::::::::::::::::::: questions 

- How do you make a Table One?
- How do you make a Table One that is easy to configure?

::::::::::::::::::::::::::::::::::::::::::::::::

::::::::::::::::::::::::::::::::::::: objectives

- Explain what a Table One is
- Know how to make a Tabel One and adjust key parameters


::::::::::::::::::::::::::::::::::::::::::::::::

## What is a "Table One"?

Primarily used in medical and epidemiological research,
a Table One is typically the first table in any publication
using data.

It presents the baseline characteristics of the participants
in a study, and provides a concise overview of the relevant
demographic and clinical variables.

It typically compares different groups (male~female, treatment~control), to highlight similarities and differences.

```{r}
1+1
```

```{r prep_table1_data, echo = FALSE, warning = FALSE, message =FALSE}
library(tidyverse)
library(table1)
blood <- read_csv("episodes/data/BLOOD.csv") %>% 
  mutate(testost = na_if(testost, 999.0)) %>% 
  mutate(prolactn = na_if(prolactn, 99.99)) %>% 
  select(-estrone)

blood$case <- factor(blood$case)
label(blood$case) <- "case_control"
levels(blood$case) <- c("control", "case")
blood$curpmh <- factor(blood$curpmh)

label(blood$curpmh) <- "current_pmh"
levels(blood$curpmh) <- c("no", "yes")
label(blood$ageblood) <- "Age"
units(blood$ageblood) <- "years"

```

It can look like this:

```{r basic-table1-demo, echo = FALSE, results='asis'}
library(gt)
table1(~ageblood+testost+prolactn|case+curpmh, data = blood) 
```


Og det gør vi så med gt i stedet.

Herunder er vi ikke helt i mål endnu. Men vi er ret tæt.

```{r}

library(dplyr)
library(gtsummary)

# grunddata
base <- blood |>
  select(age, grade, stage, trt) |>
  mutate(grade = paste("Grade", grade))

# rækkefølge på paneler: Overall først, derefter Grade I/II/III
lvl <- c("Overall", paste("Grade", levels(trial$grade)))

# lav et samlet datasæt med et ekstra "Overall"-panel
df <- bind_rows(
  base |> mutate(.panel = "Overall"),
  base |> mutate(.panel = grade)
) |>
  mutate(.panel = factor(.panel, levels = lvl))

# tabel: tre strata + et overall-stratum
tbl <- df |>
  tbl_strata(
    strata = .panel,
    .tbl_fun = ~ .x |>
      select(-grade) |> 
      tbl_summary(by = trt, missing = "no") |>
      add_n(),
    .header = "**{strata}**, N = {n}"
  )

tbl

```

:::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::: instructor

Primarily of use if there are medical students on the course

::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::


::::::::::::::::::::::::::::::::::::: keypoints 

- A Table One provides a compact describtion of the data we are working with
- With a little bit of work we can control the content of the table.

::::::::::::::::::::::::::::::::::::::::::::::::

